---
title: "{{< var study_title >}}"
subtitle: "Logistic Regression — Social Determinants of Diabetes"
---

## Setup

```{r}
#| label: setup
#| collapse: true
#| code-summary: "Load packages and helpers"

source(here::here("R", "setup.R"))
library(car)              # vif()
library(performance)      # check_model(), check_collinearity()
library(pROC)             # roc(), auc(), ggroc()
library(ResourceSelection) # hoslem.test()
```

---

## Data Import

```{r}
#| label: import

df_raw <- read_parquet(here("data", "processed", "analysis_dataset.parquet"))
cat("Rows in analytical dataset:", nrow(df_raw), "\n")
```

---

## Data Preparation

```{r}
#| label: prep-factors
#| code-summary: "Set factor levels and reference categories"

df <- df_raw |>
  mutate(
    diabetes = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),

    age_group = factor(age_group,
                       levels = c("18-34", "35-44", "45-54", "55-64", "65-74", "75+")),

    sex = factor(sex, levels = c("Male", "Female")),

    race_ethnicity = factor(race_ethnicity,
                            levels = c("NH White", "NH Black", "NH AIAN", "NH Asian",
                                       "NH NHOPI", "NH Other/Multiracial",
                                       "Hispanic", "NH Other")),

    education = factor(education,
                       levels = c("Graduated college", "Some college",
                                  "Graduated HS", "Did not graduate HS")),

    income_group = factor(income_group,
                          levels = c(">$100k", "$50k-$100k", "$35k-$50k",
                                     "$25k-$35k", "$15k-$25k", "<$15k")),

    bmi_category = factor(bmi_category,
                          levels = c("Normal", "Underweight", "Overweight", "Obese")),

    physically_active = factor(physically_active, levels = c(1, 0),
                               labels = c("Active", "Inactive")),

    smoking_status = factor(smoking_status,
                            levels = c("Never", "Former",
                                       "Current (some days)", "Current (daily)")),

    heavy_drinker = factor(heavy_drinker, levels = c(0, 1),
                           labels = c("No", "Yes")),

    has_provider = factor(has_provider, levels = c(1, 0),
                          labels = c("Has provider", "No provider")),

    mental_health_days = factor(mental_health_days,
                               levels = c("0 days", "1-13 days", "14+ days"))
  )
```

### Complete case exclusions

```{r}
#| label: complete-cases
#| collapse: true

predictor_cols <- c(
  "age_group", "sex", "race_ethnicity", "education", "income_group",
  "bmi_category", "physically_active", "smoking_status",
  "heavy_drinker", "has_provider", "mental_health_days"
)

n_before <- nrow(df)
df_cc <- df |> drop_na(all_of(predictor_cols))
n_after  <- nrow(df_cc)

cat("Before complete case exclusion:", n_before, "\n")
cat("After complete case exclusion: ", n_after,  "\n")
cat("Excluded:                      ", n_before - n_after, "\n")
cat("Retained (%):                  ", round(n_after / n_before * 100, 1), "%\n")
```

---

## Model

### Formula

```{r}
#| label: model-formula

f <- formula(
  diabetes ~
    age_group + sex + race_ethnicity +
    education + income_group + has_provider +
    bmi_category + physically_active + smoking_status +
    heavy_drinker + mental_health_days
)
f
```

### Fit

```{r}
#| label: model-fit
#| collapse: true

mod <- glm(f, data = df_cc, family = binomial(link = "logit"))
summary(mod)
```

### Multicollinearity — VIF

```{r}
#| label: model-vif
#| collapse: true

vif(mod)
```

### Diagnostics

```{r}
#| label: model-diagnostics
#| fig-width: 10
#| fig-height: 9

check_model(mod)
```

---

## Model Performance

### ROC curve and AUC

```{r}
#| label: model-roc
#| fig-cap: "ROC curve — full social determinants model"
#| fig-width: 6
#| fig-height: 6

roc_obj <- roc(
  response  = as.integer(df_cc$diabetes) - 1L,
  predictor = fitted(mod),
  quiet     = TRUE
)

ggroc(roc_obj, color = palette_epi["vermillion"], linewidth = 1) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed", color = "grey60") +
  annotate("text", x = 0.35, y = 0.1,
           label = paste0("AUC = ", round(auc(roc_obj), 3)),
           size  = 4.5) +
  labs(title = "ROC Curve — Full Model", x = "Specificity", y = "Sensitivity")
```

### Base demographic model vs full SDoH model — AUC comparison

```{r}
#| label: model-auc-comparison
#| collapse: true

mod_base <- glm(
  diabetes ~ age_group + sex + race_ethnicity,
  data   = df_cc,
  family = binomial
)

roc_base <- roc(
  response  = as.integer(df_cc$diabetes) - 1L,
  predictor = fitted(mod_base),
  quiet     = TRUE
)

cat("Base model AUC (age + sex + race): ", round(auc(roc_base), 3), "\n")
cat("Full model AUC (all predictors):   ", round(auc(roc_obj),  3), "\n")
```

### Hosmer-Lemeshow goodness of fit

```{r}
#| label: model-hosmer-lemeshow
#| collapse: true

hl <- hoslem.test(as.integer(df_cc$diabetes) - 1L, fitted(mod), g = 10)
print(hl)
# p > 0.05 indicates adequate fit
```

---

## Results

### Multivariable odds ratio table

```{r}
#| label: results-table

tbl_regression(
  mod,
  exponentiate = TRUE,
  estimate_fun = \(x) style_sigfig(x, digits = 3)
) |>
  bold_p(t = 0.05) |>
  bold_labels() |>
  add_n()
```

### Univariable ORs

```{r}
#| label: results-univariable

df_cc |>
  select(diabetes, age_group, sex, race_ethnicity, education, income_group,
         bmi_category, physically_active, smoking_status,
         heavy_drinker, has_provider, mental_health_days) |>
  tbl_uvregression(
    method       = glm,
    y            = diabetes,
    method.args  = list(family = binomial),
    exponentiate = TRUE,
    estimate_fun = \(x) style_sigfig(x, digits = 3)
  ) |>
  bold_p(t = 0.05) |>
  bold_labels()
```

### Tidy coefficients

```{r}
#| label: results-tidy
#| collapse: true

tidy(mod, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

---

## Conclusions / Notes

> *(Interpret key ORs in a public health context, note model limitations — separation in small race/ethnicity subgroups, unweighted analysis, cross-sectional design.)*

---

## Session Info

```{r}
#| label: session-info
#| collapse: true
#| code-summary: "R session info"

sessionInfo()
```
